Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-28 pnas/Towards.tex
no message

In this section, we present a toy model which is studied in the next sections. 
The role of this model is to introduce the mathematical principles enabling to estimate a signal in a low SNR regime, although its detection might be impossible. Later, we discuss how these principles can be carried through for the SPR problem. The model is formulated for 1-D signals to ease exposition.

Let $x\in\RL$ be the target signal and let $y\in\RN$ be the observed data. Let  $s \in \{0, 1\}^{N-L+1}$ be a binary signal indicating (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
While both $x$ and $s$ are unknown, the goal is only to estimate $x$ from $y$. The parameters of the signal $s$ (the locations of its nonzero values) are the \emph{nuisance variables} of the problem. As will be shown next, in the low SNR environment, estimating $s$ is impossible, where estimating $x$ is tractable under some conditions on $s$.

The binary signals obey the following property:
\begin{align} 
\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

Our problem can be interpreted as a special case of the \emph{system identification} problem. Similarly to~\eqref{eq:model}, the system identification forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the ``system''), $w$ is an unknown, random, input sequence, and $\varepsilon$ is an additive noise.   
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w$, satisfying the condition of~\eqref{eq:spacing}, we obtain our model. The same observation model is used for \emph{blind deconvolution}, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few.

@@

Under the space constraint~\eqref{eq:spacing}, the relation between autocorrelations of the micrograph and those of $x$ is particularly simple. It is useful to introduce some notation: let $M$ denote the number of occurrences of $x$ in $y$, and let
\begin{align}
\gamma & = \frac{M L}{N}
\end{align}
denote the density of $x$ in $y$ (that is, the fraction of entries of $y$ occupied by occurrences of $x$.) The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

One simple observation is that the first-order autocorrelation of $y$ (its mean) is independent of the locations of $x$. Since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:\footnote{We did not fully specify a random generating model for the location vector $s$. The expectation is still well defined specifically because the quantity under consideration is independent of $s$ under the assumptions.}
\begin{align*}
\E\{ a_y^1 \} & = \gamma a_x^1.
\end{align*}
We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density.) In that regime, the law of large numbers gives meaning to the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1.
\end{align}
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)
\TODO{TB: I think that this paragraph can be much shorter and it is sufficient to show only the second equation and remove footnote. Nicolas thinks otherwise. Need to be decided. }

The spacing constraint~\eqref{eq:spacing} gives rise to more powerful observations. Consider the second-order autocorrelation in particular: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
\begin{align} \label{eq:ac2_micrograph}
\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell]
\end{align}
for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.
Along the same lines, one can establish a relation for third-order autocorrelations:
\begin{align} \label{eq:ac3_micrograph}
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] \\& + \sigma^2\gamma a_{x}^1 \cdot \big(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]\big), 
 \nonumber
\end{align}
for $\ell_1,\ell_2 = 0, \ldots, L-1$. Here too, few entries are affected by $\sigma$ in the limit.
Detailed derivations for identities in this and the next part are given in Appendix~\ref{sec:autocorrelation_computation}.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedures, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations.
%%%and therefore the assumption that the first and last entries are nonzero is met.
In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$\approx1/375$). This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

@@

In practice, we do not expect to know $\gamma$ and maybe not even $\sigma$.
Figure~\ref{fig:1D_example} shows recovery of a 1-D signal from the first three autocorrelations of the data.
The autocorrelations are computed from noisy micrographs with $\sigma=3$ that obey the separation condition of~\eqref{eq:spacing}.
 Both the signal  and $\gamma$ are estimated simultaneously from the observed autocorrelations by LS fitting; more details are given in Appendix~\ref{sec:numeric_details}. Crucially, the LS does not require the knowledge of $\sigma$.

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an SNR of approximately $1/375$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

@@

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far until reaching the desried number of signal occurences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.
%	 fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
%This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$
%% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/9.

@@

\paragraph{3-D experiments.}
For the experiment in Figure~TKTK, we computed the analytical first three autocorrelations of the volume according to~TKTK.

@@

Let $\II\in\RNN$ denote a micrograph. We assume it consists of shifted copies of projections contaminated by additive white Gaussian noise:
\begin{equation}\label{eq:micrograph_model}
\II = \sum_{t=1}^{M} I_{\omega_t}\ast\delta_{\mb s_t}+ \varepsilon, \quad \varepsilon~\sim\mathcal{N}(0,\sigma^2 I),
\end{equation}
where the viewing directions $\omega_t$ are assumed to be drawn from the uniform distribution over $\text{SO}(3)$ and $\mb s_t$ denotes the location of the center of the $t$th projection in the micrograph. Similarly to~\eqref{eq:spacing}, we impose a separation condition so that any two projections are separated by at least $2P-1$ pixels between their upper left corners  in each direction.
Note that~\eqref{eq:model} can be also written as a sum of $\delta$ functions as in~\eqref{eq:micrograph_model}.

@@

Recall that we obtain the autocorrelations of the volume in PSWFs coefficients; see~\eqref{eq:third_order_ac_pswf}. Hence, to conclude the derivation we expand
\[ \langle a^3_{I_{\omega}}[\Bell_1, \Bell_2]\rangle_{\omega}= \sum_{k,q_1,q_2}\mathfrak{a}^3[k,q_1,q_2]\psi_{k,q_1}[\Bell_1]\overline{\psi_{k,q_2}[\Bell_2]},\]
where we only include the block-diagonal terms in the expansion; the rest are equal to zero. Let
\begin{align*}
\Psi_{\ell,N,s}[\Bell] &:= \sum_{n=0}^{n_{\text{max}}(N)}\beta_{\ell,s;N,n}\psi_{N,n}[\Bell],\\ 
\rho_{\ell,N,s}^{(k,q)}&:=\int_{\Bell}\Psi_{\ell,N,s}([\mb i+\Bell])\overline{\psi_{k,q}(\Bell)}.
\end{align*}
Then, 
the final formula reads
\[\begin{aligned} &\mathfrak{a}^3[k,q_1,q_2] = \sum_{\substack{\ell_1,m_1,s_1\\\ell_2,m_2,s_2\\s_3}}\sum_{\ell_3=|\ell_1-\ell_2|}^{\min(L,\ell_1+\ell_2)}\tamir_{\ell_1,m_1,s_1}\tamir_{\ell_2,m_2,s_2}\overline{\tamir_{\ell_3,m_1+m_2,s_3}}\\
&\times (-1)^{m_1+m_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\\ m_1 & m_2 & -m_1-m_2\end{array}\right)\\
&\times \sum_{N_1=-\ell_1}^{\ell_1}\sum_{N_2=-\ell_2}^{\ell_2}(-1)^{N_1+N_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\\ N_1 & N_2 & -N_1-N2\end{array}\right)\\&\times\frac{1}{L^2}\sum_{\mb i}\Psi_{\ell_1,N_1,s_1}[\mb i]\rho_{\ell_2,N_2,s_2}^{(k,q_1)}[\mb i]\overline{\rho_{\ell_3,N_1+N_2,s_3}^{(k,q_2)}[\mb i]}. \end{aligned}\]
In practice, the last two lines of the above expression for $\mathfrak{a}^3[k,q_1,q_2]$ is precomputed, and both the integration over $\mb i$ and over $\Bell$ is performed on the grid of the images in the dataset, to match the integration performed on the actual images.

@@

As before, we want to derive the relation with respect to the PSWF coefficients of the autocorrelation. Hence,  we expand the above in 2-D PSWFs by
\[\left\langle a^2_{\hat{I}_{\omega}}(k,\theta)\right\rangle_{\omega} =
\sum_{q}\mathfrak{a}_2(q)\psi_{0,q}(k),\]
and conclude that
\[ \mathfrak{a}^2[q] =
\frac{1}{\sqrt{8\pi}}\sum_{\substack{\ell,m\\s_1,s_2}}\tamir_{\ell,m,s_1}\overline{\tamir_{\ell,m,s_2}}
\int_0^1j_{\ell,s_1}(k)j_{\ell,s_2}(k)R_{0,q}(k)k\, dk.\]
The integral on $k$ is precomputed.

@@

%Since the Fourier transformed volume is given by
%\[ \widehat V(k,\theta,\varphi) =
%\sum_{\ell,m,s}a_{\ell,m,s}Y_{\ell,m}(\theta,\varphi)j_{\ell,s}(k),\]
Since $j_{\ell,s}(0) = 0$ unless $\ell=0$, and since
$Y_{0,0}(\theta,\varphi) = \frac{1}{\sqrt{4\pi}}$, we conclude from~\eqref{eq:volume_expansion_app} that
\[ \langle
a^1_{I_{\omega}}\rangle_{\omega} = \widehat \phi(\mb 0) = \frac{1}{\sqrt{4\pi}}\sum_s\tamir_{0,0,s}j_{0,s}(0).\]

>>>>

In this section, we present a toy model which is studied in the next sections. 
The role of this model is to introduce the mathematical principles enabling estimation of a signal in a low SNR regime, even when detection is impossible. Later, we discuss how these principles can be carried through for the SPR problem. The model is formulated for 1-D signals to ease exposition.

Let $x\in\RL$ be the target signal and let $y\in\RN$ be the observed data, where $N \gg L$. Let  $s \in \{0, 1\}^{N-L+1}$ be a binary signal indicating (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
While both $x$ and $s$ are unknown, the goal is only to estimate $x$ from $y$. The parameters of the signal $s$ (the locations of its nonzero values) are the \emph{nuisance variables} of the problem. As will be shown next, in the low SNR environment, estimating $s$ is impossible, whereas estimating $x$ is tractable under some conditions on $s$.

We assume that the binary signals $s$ obey the following \emph{separation condition}: If $s[i] = 1$ and $s[j] = 1$  for  $i \neq j$, then $\vert i - j\vert \geq 2L-1$.
%\begin{align} 
%\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
%\label{eq:spacing}
%\end{align}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

Our problem can be interpreted as a special case of the \emph{system identification} problem. Similarly to~\eqref{eq:model}, the system identification forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the ``system''), $w$ is an unknown, random, input sequence, and $\varepsilon$ is an additive noise.   
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w$, satisfying the separation condition, we obtain our model. The same observation model is used for \emph{blind deconvolution}, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few.

@@

Under the spacing constraint, the relation between autocorrelations of the micrograph and those of $x$ is particularly simple. It is useful to introduce some notation: let $M$ denote the number of occurrences of $x$ in $y$ (that is, the number of 1's in $s$), and let
\begin{align}
\gamma & = \frac{M L}{N}
\end{align}
denote the density of $x$ in $y$ (that is, the fraction of entries of $y$ occupied by occurrences of $x$.) The spacing constraint%~\eqref{eq:spacing} 
imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

One simple observation is that the first-order\TODO{Will: I suggest making this statement about all the autocorrelations here, instead of singling out the mean first. I understand that the mean can be estimated without the separation condition, but this distinction is not very critical to highlight, and it's a little confusing to have this sentence (upon first reading, it might be interpreted to mean that the other autocorrelations do depend on s). I'd say something like: Under the separation condition on $s$, the autocorrelation functions of $y$ can be shown to depend only on the corresponding autocorrelations of $x$ and the parameter $gamma$, defined by... 
Then go into the details.} autocorrelation of $y$ (its mean) is independent of the locations of $x$. Since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:\TODO{Will: Instead of the footnote, just specify here that the randomness is over the Gaussian noise.
s can be deterministic. The only thing we assume about s is that the associated parameter gamma is constant (you might also note that this can be relaxed, so long as the sequence of $\gamma_N$'s converges to gamma, but this point is not critical).}\footnote{We did not fully specify a random generating model for the location vector $s$. The expectation is still well defined specifically because the quantity under consideration is independent of $s$ under the assumptions.}
\begin{align*}
\E\{ a_y^1 \} & = \gamma a_x^1.
\end{align*}
We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density.) In that regime, the law of large numbers can be used to show the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1.
\end{align}
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)
\TODO{TB: I think that this paragraph can be much shorter and it is sufficient to show only the second equation and remove footnote. Nicolas thinks otherwise. Need to be decided. }

The spacing condition gives rise to more powerful observations. Consider the second-order autocorrelation in particular: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
\begin{align} \label{eq:ac2_micrograph}
\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell]
\end{align}
for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.
Along the same lines, one can establish a relation for third-order autocorrelations:
\begin{align} \label{eq:ac3_micrograph}
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] \\& + \sigma^2\gamma a_{x}^1 \cdot \big(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]\big), 
 \nonumber
\end{align}
for $\ell_1,\ell_2 = 0, \ldots, L-1$. Here too, few entries are affected by $\sigma$ in the limit.
Detailed derivations for identities in this and the next part are given in Appendix~\ref{sec:autocorrelation_computation}.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedures, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the separation condition holds, then the length of the signal can be determined from the autocorrelations.
%%%and therefore the assumption that the first and last entries are nonzero is met.
In particular, if the separation condition holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$\approx1/375$). This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with the separation condition. %~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

@@

In practice, we do not expect to know $\gamma$ and maybe not even $\sigma$.
Figure~\ref{fig:1D_example} shows recovery of a 1-D signal from the first three autocorrelations of the data.
The autocorrelations are computed from noisy micrographs with $\sigma=3$ that obey the separation condition.% of~\eqref{eq:spacing}.
 Both the signal  and $\gamma$ are estimated simultaneously from the observed autocorrelations by LS fitting; more details are given in Appendix~\ref{sec:numeric_details}. Crucially, the LS does not require the knowledge of $\sigma$.

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition.%~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an SNR of approximately $1/375$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

@@

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.
%	 fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
%This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$
%% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/9.

@@

\paragraph{3-D experiments.}

For the experiment in Figure~TKTK, we computed the analytical first three autocorrelations of the volume according to~TKTK. To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$.

The LS was solved by TKTK. The expected $\hat \gamma$ is one.

In the first experiment, we used \TODO{Bla bla bla, real valued volume}

The micrographs for the experiment presented in Figure~\ref{fig:cryo_detection} were generated as follows \TODO{separation, etc. }. Then, the LS~\eqref{eq:LS_cryo} was solved with $L=0$ (low-resolution estimate).

@@

Let $\II\in\RNN$ denote a micrograph. We assume it consists of shifted copies of projections contaminated by additive white Gaussian noise:
\begin{equation}\label{eq:micrograph_model}
\II = \sum_{t=1}^{M} I_{\omega_t}\ast\delta_{\mb s_t}+ \varepsilon, \quad \varepsilon~\sim\mathcal{N}(0,\sigma^2 I),
\end{equation}
where the viewing directions $\omega_t$ are assumed to be drawn from the uniform distribution over $\text{SO}(3)$ and $\mb s_t$ denotes the location of the center of the $t$th projection in the micrograph. %Similarly to~\eqref{eq:spacing}, 
Again, we impose a separation condition so that any two projections are separated by at least $2P-1$ pixels between their upper left corners  in each direction.
Note that~\eqref{eq:model} can be also written as a sum of $\delta$ functions as in~\eqref{eq:micrograph_model}.

@@

Recall that we obtain the autocorrelations of the volume in PSWFs coefficients; see~\eqref{eq:third_order_ac_pswf}. Hence, to conclude the derivation we expand
\[ \langle a^3_{I_{\omega}}[\Bell_1, \Bell_2]\rangle_{\omega}= \sum_{k,q_1,q_2}\mathfrak{a}_x^3[k,q_1,q_2]\psi_{k,q_1}[\Bell_1]\overline{\psi_{k,q_2}[\Bell_2]},\]
where we only include the block-diagonal terms in the expansion; the rest are equal to zero. Let
\begin{align*}
\Psi_{\ell,N,s}[\Bell] &:= \sum_{n=0}^{n_{\text{max}}(N)}\beta_{\ell,s;N,n}\psi_{N,n}[\Bell],\\ 
\rho_{\ell,N,s}^{(k,q)}&:=\int_{\Bell}\Psi_{\ell,N,s}([\mb i+\Bell])\overline{\psi_{k,q}(\Bell)}.
\end{align*}
Then, 
the final formula reads
\begin{align} \label{eq:ac_3_prolates}
 &\mathfrak{a}_x^3[k,q_1,q_2] = \sum_{\substack{\ell_1,m_1,s_1\nonumber \\ \ell_2,m_2,s_2\\s_3}}\sum_{\ell_3=|\ell_1-\ell_2|}^{\min(L,\ell_1+\ell_2)}\tamir_{\ell_1,m_1,s_1}\tamir_{\ell_2,m_2,s_2}\overline{\tamir_{\ell_3,m_1+m_2,s_3}}\\
&\times (-1)^{m_1+m_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\\ m_1 & m_2 & -m_1-m_2\end{array}\right)\\
&\times \sum_{N_1=-\ell_1}^{\ell_1}\sum_{N_2=-\ell_2}^{\ell_2}(-1)^{N_1+N_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\nonumber\\ N_1 & N_2 & -N_1-N2\end{array}\right)\nonumber\\&\times\frac{1}{L^2}\sum_{\mb i}\Psi_{\ell_1,N_1,s_1}[\mb i]\rho_{\ell_2,N_2,s_2}^{(k,q_1)}[\mb i]\overline{\rho_{\ell_3,N_1+N_2,s_3}^{(k,q_2)}[\mb i]}. \nonumber
\end{align}
%
%\[\begin{aligned} \label{eq:ac_3_prolates} &\mathfrak{a}_x^3[k,q_1,q_2] = \sum_{\substack{\ell_1,m_1,s_1\\\ell_2,m_2,s_2\\s_3}}\sum_{\ell_3=|\ell_1-\ell_2|}^{\min(L,\ell_1+\ell_2)}\tamir_{\ell_1,m_1,s_1}\tamir_{\ell_2,m_2,s_2}\overline{\tamir_{\ell_3,m_1+m_2,s_3}}\\
%&\times (-1)^{m_1+m_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\\ m_1 & m_2 & -m_1-m_2\end{array}\right)\\
%&\times \sum_{N_1=-\ell_1}^{\ell_1}\sum_{N_2=-\ell_2}^{\ell_2}(-1)^{N_1+N_2}\left(\begin{array}{ccc}\ell_1 & \ell_2  & \ell_3\\ N_1 & N_2 & -N_1-N2\end{array}\right)\\&\times\frac{1}{L^2}\sum_{\mb i}\Psi_{\ell_1,N_1,s_1}[\mb i]\rho_{\ell_2,N_2,s_2}^{(k,q_1)}[\mb i]\overline{\rho_{\ell_3,N_1+N_2,s_3}^{(k,q_2)}[\mb i]}. \end{aligned}\]
In practice, the last two lines of the above expression for $\mathfrak{a}_x^3[k,q_1,q_2]$ is precomputed, and both the integration over $\mb i$ and over $\Bell$ is performed on the grid of the images in the dataset, to match the integration performed on the actual images.

@@

As before, we want to derive the relation with respect to the PSWF coefficients of the autocorrelation. Hence,  we expand the above in 2-D PSWFs by
\[\left\langle a^2_{\hat{I}_{\omega}}(k,\theta)\right\rangle_{\omega} =
\sum_{q}\mathfrak{a}_x^2(q)\psi_{0,q}(k),\]
and conclude that 
\begin{equation} \label{eq:ac_2_prolates}
\mathfrak{a}_x^2[q] =
\frac{1}{\sqrt{8\pi}}\sum_{\substack{\ell,m\\s_1,s_2}}\tamir_{\ell,m,s_1}\overline{\tamir_{\ell,m,s_2}}
\int_0^1j_{\ell,s_1}(k)j_{\ell,s_2}(k)R_{0,q}(k)k\, dk.
\end{equation}
The integral on $k$ is precomputed.

@@

%Since the Fourier transformed volume is given by
%\[ \widehat V(k,\theta,\varphi) =
%\sum_{\ell,m,s}a_{\ell,m,s}Y_{\ell,m}(\theta,\varphi)j_{\ell,s}(k),\]
Since $j_{\ell,s}(0) = 0$ unless $\ell=0$, and since
$Y_{0,0}(\theta,\varphi) = \frac{1}{\sqrt{4\pi}}$, we conclude from~\eqref{eq:volume_expansion_app} that
\begin{equation} \label{eq:ac_1_prolates}
\mathfrak{a}_x^1 = \langle
a^1_{I_{\omega}}\rangle_{\omega}  = \widehat  \phi(\mb 0) = \frac{1}{\sqrt{4\pi}}\sum_s\tamir_{0,0,s}j_{0,s}(0).
\end{equation}
